---
title: "Sorting by fractal dimension"
author: "Steve Simon"
date: "March 12, 2018"
output: html_document
---

WARNING!! The geom_sf function in ggplot2 is experimental and not yet part of the regular package (at least not at the time I wrote this, February 2018). You have to install a special version of ggplot2 using the RTools package. For details, go to

https://dcl-2017-01.github.io/curriculum/notes/spatial-vis.html

and note the installation instructions at the very top of the page.

```{r prelims}
library(ggmap)
library(ggplot2)
library(geosphere)
library(magrittr)
library(maptools)
library(rvest)
library(sf)
library(tidyverse)
library(viridis)
```

The fractal dimension was estimated by box counting and the results stored in csv files in the results subdirectory.

```{r read-box-counts}
pn <- "~/mapping-shapefiles/results"
pa <- "*.csv"
fn <- list.files(pn, pa)
n <- length(fn)
fd <- data.frame(district=rep(-1, n), fractal_dimension=rep(-1, n))
for (i in 1:n) {
  fr <- read.csv(paste(pn, fn[i], sep="/"), header=TRUE)
  fd$district[i] <- fr$d[1]
  fd$fractal_dimension[i] <- coef(lm(log2(y)~x, data=fr))[2]
}
fd$district <- paste0(ifelse(fd$district <= 999, "0", ""), fd$district) 
fd %<>% arrange(desc(fractal_dimension))
head(fd, 20)
```

The file used in this example, cb_2016_us_cd115_500k.shp, was downloaded from https://www.census.gov/geo/maps-data/data/cbf/cbf_cds.html.

The 500K means that the scale is 1:500,000 which is the highest resolution available.

```{r read-shapefiles}
path_name <- "~/mapping-shapefiles/shp"
file_name <- "cb_2016_us_cd115_500k.shp"
path_name                              %>%
  paste(file_name, sep="/")            %>%
  st_read(stringsAsFactors=FALSE)      -> all_districts
```

Mapping functions

```{r mapping}
highlight_district <- function(district_number) {
  state_fips_code <- substr(district_number, 1, 2)
  district_string <- as.character(district_number)
  all_districts                          %>%
    mutate(fill_color=
      ifelse(GEOID==district_number,
        district_string, "others"))      -> single_district
  single_district                        %>%
    filter(STATEFP==state_fips_code)     %>%
    ggplot()                              +
    geom_sf(aes(fill=fill_color))         +
    scale_fill_manual(
      values=c("gray", "white"))          +
    theme(legend.position="none")
}
isolate_district <- function(district_number) {
  district_string <- as.character(district_number)
  all_districts                          %>%
    filter(GEOID==district_string)       %>%
    ggplot()                              +
    geom_sf(fill="gray")       
}
```

```{r map-districts, fig.width=4, fig.height=4}
for (i in 1:20) {
  fd$district[i]                         %>%
    highlight_district                   %>%
    plot
  fd$district[i]                         %>%
    isolate_district                     %>%
    plot
}
```

```{r save-everything}
save.image("~/mapping-shapefiles/data/fractal01.RData")
```
